---
title: "Assignment 1 in Reproducible Research"
author: "Carla Wright"
date: "Thursday, March 12, 2015"
output: md_document
---
##Loading and preprocessing the data
1. Load the data (i.e. read.csv())
```{r}
if (!file.exists("activity.csv")) {
                download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", destfile = "activity.zip", mode="wb")
                unzip("activity.zip")
                unlink("activity.zip")
        write(date(), file="dateDownloaded.txt")
}
activity<-read.csv("activity.csv")
```

2. Process/transform the data (if necessary) into a format suitable for your analysis
Remove NA values
```{r}
cleanactivity <- na.omit(activity)
```

##Question 1:What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day
```{r}
stepsbydate <- aggregate(steps ~ date, cleanactivity, sum)
```

2. Make a histogram of the total number of steps taken each day
```{r, echo=TRUE}
hist(stepsbydate$steps, col=1, main="Total Steps per Day", 
     xlab="Total number of steps per day")
```

3. Calculate and report the mean and median of the total number of steps taken per day

```{r}
stepsmean <- mean(stepsbydate$steps)
stepsmedian <- median(stepsbydate$steps)
```
The mean and median total number of steps per day are 10766 and 10765 steps respectively.

##Question 2: What is the average daily activity pattern?

1.Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, average

Aggregate steps on each interval to get average number of steps in an interval across all days
```{r, echo=TRUE}
stepsbyinterval <- aggregate(steps ~ interval, cleanactivity, mean)

plot(stepsbyinterval$interval, stepsbyinterval$steps, type='l', col=11, 
     main="Average number of steps averaged over all days", xlab="Interval", 
     ylab="Average number of steps")
```

2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

Find row id with the maximum average number of steps in an interval
```{r}
maxstepsid <- which.max(stepsbyinterval$steps)
```
Get the interval with maximum average number of steps in an interval
```{r}
stepsbyinterval [maxstepsid , ]
```
Table id 104 which is interval 835 has the maximum average number of steps (206.2).

## Question 3:Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).
```{r}
sum(as.numeric(is.na(activity$steps)))
```
2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

To fill in the missing values of the dataset, my approach is to replace the NA by the mean for that 5-minute interval. This data has already been calculated in the data frame "stepsbyinterval".


3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

The new dataset is called "newvalues"
```{r}
newvalues <- transform(activity, steps = ifelse(is.na(activity$steps), 
                                        stepsbyinterval$steps[match(activity$interval, 
                                        stepsbyinterval$interval)], activity$steps))
```

Aggregate steps to get total number of steps in a day
```{r}
totalstepbydate <- aggregate(steps ~ date, newvalues, sum)
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 
```{r, echo=TRUE}
hist(totalstepbydate$steps, col=1, main="Total Steps per Day", 
     xlab="Total number of steps per day")
```


Do these values differ from the estimates from the first part of the assignment?
What is the impact of imputing missing data on the estimates of the total daily number of steps?

The mean and median total number of steps per day are both 10766 respectively. The results which include replacement values(the average steps within interval) of missing values slightly differ from the median results that exlcude missing values by approximately 0.01%.

##Question 4: Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

Convert date formats
```{r}
newvalues$date <- as.Date(newvalues$date, "%Y-%m-%d")
```
Add a new column indicating day of the week 
```{r}
newvalues$day <- weekdays(newvalues$date)
```
Add a new column called day type and initialize to weekday
```{r}
newvalues$typeofday <- c("weekday")
```
If day is Saturday or Sunday, write typeofday as weekend
```{r}
for (i in 1:nrow(newvalues)){
        if (newvalues$day[i] == "Saturday" || newvalues$day[i] == "Sunday"){
                newvalues$typeofday[i] <- "weekend"
        }
}
```
Convert typeofday in to factor for plotting
```{r}
newvalues$typeofday <- as.factor(newvalues$typeofday)
```
Aggregate steps as interval to get average number of steps in an interval across all days
```{r}
stepsbyday <- aggregate(steps ~ interval+typeofday, newvalues, mean)
```
2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
Load lattice for plotting
```{r, echo=TRUE}
library(lattice)

xyplot(stepsbyday$steps ~ stepsbyday$interval|stepsbyday$typeofday, main="Average Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")
```